---
title: "Spatial Operations"
author: "R.J. Dyer"
date: "September 19, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_knit$set(global.par = TRUE)
```

```{r echo=FALSE}
par( bg=NA )
```


# Raster & Vector Operations


## Common Workflow

Many analyses that integrate raster & vector data proceed along the following workflow plan:

1. Attain raster & point/line data
2. Modify raster for suitability (reprojection, cropping, etc).
3. Extract data from raster along points/lines
4. Analyze ...


## Cropping Rasters



## Extracting Point Data



## Extracting Line Data



# Habitat Specification




# Data Persistence

## Saving & Exporting Vector & Raster Objects

As all of these objects are R objects, they can be saved to disk using the `save()` function, which makes them a *.rda object.  If you have objects that take a bit of time to create, it is in your best interests to save them after creation and on subsequent analyses, just use the saved versions.

There are many situations where you need to save a raster you've manipulated.  As these raster objects are R objects, you can save them directly to the file system using the `write()` as:

```{r eval=FALSE}
write( baja_california, filename="baja_california.rda")
```

This will save the raster object to file exactly like it is in your R session.  To load it back in you just use `load()` and it is returned just like it was.  The benefit to saving these as R objects is that you do not need to change it at all to pick up where you left off.



You may also need to export the vector or raster data into a non-R format for external analyses.  A common format for vector data is the ubiquitous ESRI shapefile.  

```{r eval=FALSE}
library(rgdal)
writeOGR( obj=SPDF, dsn="spatial_data", layer="SpatialPolyExample", driver="ESRI Shapefile")
```


For rasters, you use the `writeRaster()` function.  The file extension is used to determine the file format used and R saves it automatically.

```{r eval=FALSE}
writeRaster( baja_california, filename="baja_california.tif")
```


