---
title: "Untitled"
author: "R.J. Dyer"
date: "September 16, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Characteristic Equation

Landscape genetics has been described as a combination of landscape ecology and population genetics.

\[
G \propto f(E)
\]


$G$ - A measure of the spatial distribution of genetic variance.  
$E$ - Variables external to the individuals/populations being examined.  
$f()$ - Mapping and relationship among $E$ to maximally explain $G$.  




## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

